---
title: "Bulk RNAseq report"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Gianluca Alessio Mariani"

params:
  sample_info_file: "./data/sample_list_variables.csv"
  utils_file: "./src/__utils_rna_seq_functions.R"
  dds_file: "./data/deseq2.dds.RData"
  ribosomal_genes_file: "./src/genes2filter_symbol.csv"
  min_sample: 10
  logfc: 1
  qvalue: 0.05
  genelists: 
    - "./data/251015_genelist_frascolla.csv"
  gene_name_type: "SYMBOL"
  design_vars: "base_condition"
  contrast_variable: "base_condition"
  e_var_tag: 
    - "shIFI6_1"
    - "shIFI6_2"
  c_var_tag: 
    - "shSCR"
---

```{r}
knitr::opts_chunk$set(echo       = FALSE,
                      message    = FALSE,
                      warning    = FALSE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      fig.align  = 'center',
                      fig.width  = 10,
                      fig.height = 8)
```

```{r parameters}
min_sample = params$min_sample
# logfc=params$logfc
# qvalue=params$qvalue
rdata_file=params$dds_file
```

## Introduction

FRASCOLLA AGGIUNGI

### Overview of the analysis steps

1) Quality control of raw data with [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
2) Adapter and quality trimming with [Trim Galore!](https://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)
3) Read alignment with [STAR](https://github.com/alexdobin/STAR).
4) Estimation of transcript and gene expression with [Salmon](https://combine-lab.github.io/salmon/).
5) Differential gene expression analysis with [DESeq2](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) (v1.49.1)
6) Gene set enrichment analysis with [ClusterProfiler](https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html) (v4.17.0)

The steps from 1 to 4 have been performed using [nf-core/rnaseq](https://nf-co.re/rnaseq) v3.18.0.
This report includes analysis from step 5 and 6

```{r libraries}
source(params$utils_file)

# LIBRARY
if (!require("DT")) install.packages('DT')
if (!require("git2r")) install.packages("git2r")
if (!require("reshape2")) install.packages("reshape2")

library(stringr)
library(glue)
library("DESeq2")
library(rtracklayer)
library(ComplexHeatmap)
library(RColorBrewer)
library(gplots) 
library(ggplot2)
library("DT")
library(reshape)
library(plotly)
library(clusterProfiler)
library(dplyr)
library(tidyr)
library(WGCNA)
library(gridExtra)
library("git2r")
library(limma)
library(tibble)
library(AnnotationDbi)
library("org.Hs.eg.db")
library(msigdbr)
library(ReactomePA)
```

## Samples Datatable

Below we present the table containing all the samples analyzed in this report.  

Each sample is described by:

  - sample: sample name defined a priori in the previous step of RNA seq analysis

  - condition_single: list of samples divided into groups based on the base condition and the time
  
  - condition_pooled: list of samples divided into groups based on the base condition and the time and grouping together the samples from the two shIFI6 conditions

  - base_condition: the baseline condition (control, knockdown, treated, etc...) irrespective of any other experimental variable/parameter
  
  - time: day after treatment

```{r load_samples, results="asis"}

# Get information
sample_info <- read.csv(params$sample_info_file, header = TRUE)
rownames(sample_info) <- sample_info$sample

# Convert to factors
sample_info[] <- lapply(sample_info, as.factor)

datatable(
   sample_info,
   extensions = 'Buttons',   
   options = list(     
        dom = 'Blfrtip',     
        buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),     
        lengthMenu = list(c(10, 25, 50, -1),                       
        c(10, 25, 50, "All"))   
        ),   
    rownames = FALSE )
```

```{r load_data_deseq, results="asis"}
# Load DESeq2 object containing raw count data
dds <- myload_filter_ribo(rdata_file,
  filter_ribo = TRUE, 
  ribo_file = params$ribosomal_genes_file
)

# Clean names
sample_info$sample <- make.names(sample_info$sample)
rownames(sample_info) <- sample_info$sample

# Get common samples
common_samples <- intersect(colnames(dds), sample_info$sample)
dds <- dds[, common_samples]
sample_info <- sample_info[common_samples, ]

# sanity check
stopifnot(all(colnames(dds) == rownames(sample_info)))

# verify alignment
stopifnot(all(colnames(dds) == rownames(sample_info)))

# Ensure sample names are aligned
rownames(colData(dds)) <- sample_info$sample
dds$sample <- sample_info$sample

design_vars <- c(params$design_vars)
# check which variables are constant
constant_vars <- check_constant_vars(sample_info, design_vars)
if (length(constant_vars) > 0) {
  message("Found constant variables: ", paste(constant_vars, collapse = ", "), ", removed from the design.")
}

# valid (non constant) variables
valid_vars <- setdiff(design_vars, constant_vars)

# Remove missing design variables
missing_vars = valid_vars[!(valid_vars %in% colnames(sample_info))]
if (length(missing_vars) > 0) {
  message("IMPORTANT!!!! Passed but missing variable removed: ", paste(missing_vars, collapse = ", "))
}
valid_vars <- setdiff(valid_vars, missing_vars)

if (length(valid_vars) == 0) {
  stop("No suitable variable for the design. Check the data.")
}
# build dynamic design formula
design_formula <- as.formula(
  paste("~", paste(valid_vars, collapse = " + "))
)

contrast_variable <- params$contrast_variable
```

```{r deseq2_allsamples}
counts <- assay(dds)
min_reads <- round(ncol(counts) / 2)
min_sample <- length(unique(sample_info[[contrast_variable]]))

out_deseq2 <- mydeseq2(
  counts = counts,
  min_reads = min_reads,
  sample_info = sample_info,
  min_sample = min_sample,
  design_formula = design_formula,
  gene_name_type = params$gene_name_type,
  contrast_variable = params$contrast_variable
)
  
genet <- my_genetable(
  res = out_deseq2$res, 
  title_1vs2 = "All_genes", 
  qvalue = params$qvalue, 
  logfc = params$logfc,
  path_option = params$path_option
)

allgenes <- genet[["res"]]
deg <- genet[["deg"]]
```

```{r deseq2_forupdown_inheatmaps}
dds_heat <- DESeqDataSetFromMatrix(countData = counts,
                              colData = sample_info,
                              design = ~ base_condition + time)
dds_heat <- DESeq(dds_heat)

res_IFI6_1_vs_SCR <- results(dds_heat, contrast = c(contrast_variable, params$e_var_tag[[1]], params$c_var_tag[[1]]))
res_IFI6_2_vs_SCR <- results(dds_heat, contrast = c(contrast_variable, params$e_var_tag[[2]], params$c_var_tag[[1]]))

res_IFI6_1_vs_SCR_filtered <- na.omit(res_IFI6_1_vs_SCR)
res_IFI6_2_vs_SCR_filtered <- na.omit(res_IFI6_2_vs_SCR)

# Converti in data frame per sicurezza
df1 <- as.data.frame(res_IFI6_1_vs_SCR_filtered)
df2 <- as.data.frame(res_IFI6_2_vs_SCR_filtered)

# Aggiungi una colonna con i nomi dei geni
df1$gene <- rownames(df1)
df2$gene <- rownames(df2)

# Generate df_combined for the heatmaps
df1$origin <- "IFI6_1_vs_SCR"
df2$origin <- "IFI6_2_vs_SCR"
# Unisci i due dataframe
df_combined <- rbind(df1, df2)
# Ordina per gene
df_combined <- df_combined[order(df_combined$gene), ]

# Merge delle due tabelle su "gene"
merged <- merge(df1[, c("gene", "log2FoldChange")],
                df2[, c("gene", "log2FoldChange")],
                by = "gene", suffixes = c("_IFI6_1_vs_SCR", "_IFI6_2_vs_SCR"))

# Funzione per calcolare il log2FC combinato
merged$log2FC_combined <- apply(merged[, c("log2FoldChange_IFI6_1_vs_SCR", "log2FoldChange_IFI6_2_vs_SCR")], 1, function(x) {
  fc1 <- x[1]
  fc2 <- x[2]
  
  # Caso 1: entrambi negativi
  if (fc1 < 0 & fc2 < 0) {
    return(min(fc1, fc2))  # più negativo
  }
  # Caso 2: entrambi positivi
  else if (fc1 > 0 & fc2 > 0) {
    return(max(fc1, fc2))  # più positivo
  }
  # Caso 3: segni diversi
  else {
    return(mean(c(fc1, fc2)))
  }
})

# Ordina in base al nuovo log2FC combinato
merged_sorted <- merged[order(merged$log2FC_combined, decreasing = TRUE), ]
```

## PCA Analysis

For the PCA (principal component analysis) and correlation analyses, gene expression data were normalized using the variance stabilizing transformation (VST) method implemented in DESeq2.

Below, we present the PCA performed on the complete set of samples. PCA was used to explore global variance in gene expression profiles across all samples.

The primary objectives of this analysis are to:

- Assess sample quality

- Determine whether samples cluster according to experimental conditions, suggesting biologically meaningful variation

- Identify potential outliers

- Detect batch effects or other sources of unwanted variation

By reducing the high-dimensional gene expression data into a few principal components, PCA provides a visual summary of the dataset’s structure.

```{r PCA_all, results="asis"}

out <- preprocessing_PCA(counts, 
    min_reads, 
    sample_info, 
    normalization = "vst",
    min_sample = 5, 
    design_formula = design_formula,
    contrast_variable = params$contrast_variable
    )

rld <- out[["rld"]]

colData(rld)$sample <- sample_info$sample
colData(rld)[[params$contrast_variable]] <- sample_info[[contrast_variable]]

p <- mypcaAnalysis_test(
  title_1vs2 = "Analysis on all samples", 
  rld = rld,
  intgroup1 = "base_condition"
)

p
```

```{r batch_effect, results="asis", echo=FALSE}

# n_batches <- length(unique(sample_info$batch))
# var_threshold <- 0.1
# pval_threshold <- 0.05

# do_batch_correction <- FALSE  # default: no correction

# if (n_batches > 1) {
#   check <- check_batch_effect(rld, batch_var = "batch", cond_var = "condition_description", var_threshold = var_threshold)
#   confound_check <- check_batch_condition_confounding(sample_info, batch_var = "batch", cond_var = "condition_description", pval_threshold = pval_threshold)
  
#   batch_var_exp_pct <- round(check$batch_var_exp * 100, 2)
#   threshold_pct <- var_threshold * 100
  
#   warning_msg <- ""
#   if (!is.null(confound_check$warning)) {
#     warning_msg <- paste0(confound_check$warning, "<br><br>")
#   }
  
#   if (confound_check$confounded) {
#     batch_text <- paste0(
#       "Samples come from ", n_batches, " different batches.<br><br>",
#       "The independence test between batch and condition indicates a strong association (p-value = ",
#       signif(confound_check$p_value, 3), ").<br>",
#       "This suggests that batch and condition are confounded, and batch correction might remove true biological signal.<br><br>",
#       "Therefore, batch correction <b>IS DISCOURAGED</b> to avoid introducing artifacts."
#     )
#     do_batch_correction <- FALSE
    
#   } else if (check$batch_significant) {
#     batch_text <- paste0(
#       "Samples come from ", n_batches, " different batches.<br>",
#       "The variance explained by batch in the first two principal components is approximately <b>", batch_var_exp_pct, "%</b>, exceeding the threshold of <b>", threshold_pct, "%</b>.<br>",
#       "Batch correction <b>IS SUGGESTED</b>."
#     )
#     do_batch_correction <- TRUE
    
#   } else {
#     batch_text <- paste0(
#       "Samples come from ", n_batches, " batches.<br>",
#       "The variance explained by batch is low (", batch_var_exp_pct, "%), below the threshold of ", threshold_pct, "%.<br>",
#       "Batch correction <b>IS DISCOURAGED</b> to preserve true biological signal."
#     )
#     do_batch_correction <- FALSE
#   }
  
#   # Print output only if batch correction is suggested or there is a warning
#   if (do_batch_correction || warning_msg != "") {
#     cat("### Batch Effect Correction\n\n")
#     cat(batch_text)
#     cat("\n\n")
#     cat(warning_msg)
#   }
  
# } else {
#   # Only one batch, no batch effect, no output
#   do_batch_correction <- FALSE
# }

```

<div style="color:red;">

### Interpretation PCA Analysis and evaluation of batch effect

Un bordello

</div>

## Correlation analysis

The Spearman correlation heatmap provides a global view of the similarity between gene expression profiles across all samples.
We calculated the pairwise Spearman correlation coefficients between samples and visualized them in a heatmap.
Rows and columns are hierarchically clustered based on these correlations to reveal patterns of similarity and potential groupings among samples.

```{r heatmap_metadata}
dds_current <- out_deseq2$dds_updated
sample_conditions <- as.data.frame(colData(dds_current)[, params$contrast_variable, drop = FALSE])
colnames(sample_conditions) <- params$contrast_variable

# Unique conditions
conditions <- unique(as.character(colData(dds_current)[[params$contrast_variable]]))
n_conditions <- length(conditions)

# Color palette
palette_colors <- brewer.pal(min(n_conditions, 8), "Set2")
palette_colors <- palette_colors[seq_len(n_conditions)]

if (n_conditions > length(palette_colors)) {
  palette_colors <- colorRampPalette(brewer.pal(8, "Set2"))(n_conditions)
}

# color mapping
annotation_colors <- list(
  Condition = setNames(palette_colors, conditions)
)
```

```{r heatmap_correlation, results="asis",fig.height=12, fig.width=14}

# Correlation between samples
cor_matrix <- cor(assay(rld), method = "spearman")

# Distance matrix defined as 1-correlation
clustering_dist <- as.dist(1 - cor_matrix)
```

### **Spearman Correlation: General Interpretation**

####  **High Correlation Values** (> 0.98)  
*Between replicates of the same condition:*  

  - A very good quality signal  

  - Indicates that replicates behave consistently  

  - Suggests well-defined and reproducible biological conditions  

*Between different conditions:*  

  - May indicate minor transcriptional differences  

  - Or poor separation due to contamination or mislabeling

#### **Lower Correlation Values** (< 0.95)  
*Between replicates of the same condition:* 

  - May suggest technical or biological issues:  

    - Library prep/sequencing errors  

    - Sample mix-up or mislabeling  

    - Biological heterogeneity  

In some cases, biological replicates may exhibit a certain degree of variability that cannot be entirely avoided. 
This is particularly true when samples are obtained from different individuals, such as patient-derived samples, even when all other experimental conditions are carefully controlled.

Therefore, lower correlation values between replicates should not be interpreted in a standardized way, but rather evaluated in the specific biological and experimental context of the study.

*Between different conditions:*  

  - Expected when conditions are biologically distinct  

  - If correlations are too similar to replicates, it may suggest:  

    - Weak treatment effects  

    - Few genes affected by the condition  


### Spearman Correlation Heatmap

Below we present the heatmap associated spearman correlation.

```{r heatmap_correlation_plot, fig.width=12, fig.height=12}

pheatmap(
  cor_matrix,
  clustering_distance_rows = clustering_dist,
  clustering_distance_cols = clustering_dist,
  annotation_col = sample_conditions,
  annotation_colors = annotation_colors,
  main = "Sample-to-sample correlation (Spearman)",
  display_numbers = TRUE,
  number_format = "%.2f",
  fontsize_number = 7,
  cellheight = 14,
  cellwidth = 14
)
```

<div style="color:red;">

##### Interpretation Spearman Correlation Heatmap

ongoing

</div>

## Gene expression Heatmaps

Below, we present some heatmaps depicting the expression levels of the selected set of genes, filtered to include only those identified as differentially expressed (DEGs). 
Consequently, the heatmap features genes from the selected list that overlap with the DEG subset. 
Accompanying the heatmap is a table summarizing the differential expression analysis results for these genes. 
Each gene is represented by multiple rows in the table, corresponding to the number of comparisons conducted.

The genes in the heatmaps are ordered specifically to highlight the differential expression between the two considered conditions: shIFI6 and shSCR.
This is accomplished by ordering the results of the differential gene expression between the two conditions by log2 fold change and:

  - taking the highest value if the log2 fold change is positive
  
  - taking the lowest value if the log2 fold change is negative
  
  - taking the average if the log2 fold change in the two cases is one negative and one positive

```{r calculate_regulation_score, results='asis'}
gene_expression <- assay(rld)
gene_expression_raw_counts <- dds_current@assays@data$counts

expr_long <- as.data.frame(gene_expression)
expr_long$gene <- rownames(expr_long)

expr_long <- pivot_longer(expr_long, 
                          cols = -gene,
                          names_to = "sample", 
                          values_to = "expression")

# 2. Unisci con la tabella sample_info per aggiungere tempo e condizione
expr_long <- left_join(expr_long, sample_info, by = "sample")

# 3. Calcola la media per gene per ogni timepoint e condizione
expr_mean <- expr_long %>%
  group_by(gene, base_condition, time) %>%
  summarise(mean_expr = mean(expression), .groups = "drop")

# 4. Calcola la pendenza tra timepoint consecutivi per ogni gene e condizione
expr_mean <- expr_mean %>%
  arrange(gene, base_condition, time) %>%
  group_by(gene, base_condition) %>%
  mutate(slope = c(diff(mean_expr), NA)) %>%  # ultima è NA
  ungroup()

# 5. Rimuovi le righe con slope NA
expr_slopes <- expr_mean %>% filter(!is.na(slope))

# 6. Calcola i punteggi per gene
regulation_score <- expr_slopes %>%
  group_by(gene) %>%
  summarise(
    reg_score = sum(abs(slope))
  ) %>%
  ungroup()
```

```{r generate_4_families_genelists}
gene_symbols <- rownames(dds)
# gene_symbols <- gene_symbols[gene_symbols %in% rownames(gene_expression)]
gene_df <- bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
go_annot <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = gene_df$ENTREZID,
  columns = c("GO", "ONTOLOGY"),
  keytype = "ENTREZID"
) %>% 
  filter(!is.na(GO))

go_bp <- go_annot %>% filter(ONTOLOGY == "BP") %>%
  group_by(ENTREZID) %>%
  summarise(GO_BP = paste(unique(GO), collapse = "; "))
go_mf <- go_annot %>% filter(ONTOLOGY == "MF") %>%
  group_by(ENTREZID) %>%
  summarise(GO_MF = paste(unique(GO), collapse = "; "))
go_cc <- go_annot %>% filter(ONTOLOGY == "CC") %>%
  group_by(ENTREZID) %>%
  summarise(GO_CC = paste(unique(GO), collapse = "; "))

msigdb_hallmark <- msigdbr(species = "Homo sapiens", category = "H")
hallmark_annot <- msigdb_hallmark %>%
  filter(entrez_gene %in% gene_df$ENTREZID) %>%
  group_by(entrez_gene) %>%
  summarise(HALLMARK_Pathways = paste(unique(gs_name), collapse = "; "))
colnames(hallmark_annot)[1] <- "ENTREZID"

custom_df <- gene_df %>%
  left_join(go_bp, by = "ENTREZID") %>%
  left_join(go_mf, by = "ENTREZID") %>%
  left_join(go_cc, by = "ENTREZID") %>%
  left_join(hallmark_annot, by = "ENTREZID") %>%
  dplyr::select(SYMBOL, ENTREZID, GO_BP, GO_MF, GO_CC, HALLMARK_Pathways)

# Hallmark Pathways
hallmark_ids <- c(
  "HALLMARK_IL2_STAT5_SIGNALING",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
)

# GO Terms - Cytokine-Related
go_cytokine_ids <- c(
  "GO:0019221",
  "GO:0001816",
  "GO:0001818",
  "GO:0001819",
  "GO:0001817",
  "GO:0034097",
  "GO:0071345",
  "GO:0061082",
  "GO:0010934",
  "GO:0010935",
  "GO:0010936"
)

# GO Terms - Inflammation-Related
go_inflammation_ids <- c(
  "GO:0006954",
  "GO:0002526",
  "GO:0002438",
  "GO:0002675",
  "GO:0050729",
  "GO:0050727"
)

# GO Terms - Interferon-Related
go_interferon_ids <- c(
  "GO:0032609",
  "GO:0032649",
  "GO:0032689"
)

filter_genes_by_terms <- function(df, columns_to_check, terms_list) {
  # columns_to_check: vettore con i nomi delle colonne in cui cercare i termini (es: c("GO_BP", "GO_MF", "GO_CC"))
  
  df %>%
    rowwise() %>%
    mutate(
      # per ogni colonna, splitta i termini in lista e controlla se c'è almeno un termine matching
      match = any(sapply(columns_to_check, function(col) {
        terms_in_col <- str_split(get(col), pattern = ";\\s*")[[1]]
        any(terms_in_col %in% terms_list)
      }))
    ) %>%
    ungroup() %>%
    filter(match) %>%
    pull(SYMBOL) %>%
    base::unique()
}

go_cols <- c("GO_BP", "GO_MF", "GO_CC")

genes_hallmark <- custom_df %>%
  rowwise() %>%
  mutate(
    terms = list(str_split(HALLMARK_Pathways, pattern = ";\\s*")[[1]]),
    match = any(terms %in% hallmark_ids)
  ) %>%
  ungroup() %>%
  filter(match) %>%
  pull(SYMBOL) %>%
  base::unique()

genes_go_cytokine <- filter_genes_by_terms(custom_df, go_cols, go_cytokine_ids)
genes_go_inflammation <- filter_genes_by_terms(custom_df, go_cols, go_inflammation_ids)
genes_go_interferon <- filter_genes_by_terms(custom_df, go_cols, go_interferon_ids)




# Lista ordinata e df originale con valori
ordered_genes <- merged_sorted$gene
# Funzione che filtra e restituisce df con gene e log2FC_combined mantenendo l'ordine
filter_and_order_df <- function(gene_list, merged_df) {
  # Filtra merged_df per geni in gene_list
  filtered_df <- merged_df[merged_df$gene %in% gene_list, c("gene", "log2FC_combined")]
  # Ordina filtered_df secondo l'ordine di merged_df (ovvero merged_sorted)
  filtered_df <- filtered_df[match(merged_df$gene, filtered_df$gene), ]
  # Rimuove eventuali NA (casi dove gene non era nella lista)
  filtered_df <- filtered_df[!is.na(filtered_df$gene), ]
  
  return(filtered_df)
}

# Applica la funzione a ciascuna delle tue liste
df_gene_symbols        <- filter_and_order_df(gene_symbols, merged_sorted)
df_genes_hallmark      <- filter_and_order_df(genes_hallmark, merged_sorted)
df_genes_go_cytokine   <- filter_and_order_df(genes_go_cytokine, merged_sorted)
df_genes_go_inflammation <- filter_and_order_df(genes_go_inflammation, merged_sorted)
df_genes_go_interferon <- filter_and_order_df(genes_go_interferon, merged_sorted)

len_hallmark <- nrow(df_genes_hallmark)
len_cytokine <- nrow(df_genes_go_cytokine)
len_inflammation <- nrow(df_genes_go_inflammation)
len_interferon <- nrow(df_genes_go_interferon)
```

```{r generate_z_matrices, results='asis'}
list_genes <- read.csv(params$genelists[[1]])[[1]]

top_n <- 50
sample_df <- sample_info
sample_df_by_time <- sample_df[order(sample_df$time, sample_df$base_condition), ]
sample_df_by_cond <- sample_df[order(sample_df$base_condition, sample_df$time), ]
z_matrix <- t(scale(t(gene_expression)))
z_matrix_list <- z_matrix[list_genes, ]

# Filter only list_genes

regulation_score_list <- regulation_score %>%
  filter(gene %in% list_genes)
df_gene_symbols_list <- df_gene_symbols[df_gene_symbols$gene %in% list_genes, ]
len_initial_list <- nrow(df_gene_symbols_list)

# MOSTREG

all_most_regulated <- regulation_score_list %>% arrange(desc(reg_score))
top_most_regulated <- regulation_score_list %>% arrange(desc(reg_score)) %>% slice(1:top_n)
all_genes_most_reg <- all_most_regulated$gene
top_genes_most_reg <- top_most_regulated$gene

genes_in_z_mostreg <- head(intersect(all_genes_most_reg, rownames(z_matrix)), 1000)
top_genes_in_z_mostreg <- intersect(top_genes_most_reg, rownames(z_matrix))

z_matrix_top <- z_matrix[top_genes_in_z_mostreg, ]
z_matrix_top_ordered_by_day <- z_matrix_top[, sample_df_by_time$sample, drop = FALSE]
z_matrix_top_ordered_by_treatment <- z_matrix_top[, sample_df_by_cond$sample, drop = FALSE]

# UP & DOWN

all_genes_up <- df_gene_symbols_list[df_gene_symbols_list$log2FC_combined > 0, ]
all_genes_up <- all_genes_up[order(all_genes_up$log2FC_combined, decreasing = TRUE), ]$gene
all_genes_down <- df_gene_symbols_list[df_gene_symbols_list$log2FC_combined < 0, ]
all_genes_down <- all_genes_down[order(all_genes_down$log2FC_combined, decreasing = FALSE), ]$gene

top_genes_up <- head(all_genes_up, top_n)
top_genes_down <- head(all_genes_down, top_n)
top_genes_in_z_up <- intersect(top_genes_up, rownames(z_matrix))
top_genes_in_z_down <- intersect(top_genes_down, rownames(z_matrix))

z_matrix_topup <- z_matrix[top_genes_in_z_up, ]
z_matrix_topdown <- z_matrix[top_genes_in_z_down, ]

z_matrix_top_ordered_by_day_up <- z_matrix_topup[, sample_df_by_time$sample, drop = FALSE]
z_matrix_top_ordered_by_day_down <- z_matrix_topdown[, sample_df_by_time$sample, drop = FALSE]

z_matrix_top_ordered_by_treatment_up <- z_matrix_topup[, sample_df_by_cond$sample, drop = FALSE]
z_matrix_top_ordered_by_treatment_down <- z_matrix_topdown[, sample_df_by_cond$sample, drop = FALSE]

cellheight <- 12  # pixel per gene (riga)
cellwidth <- 12   # pixel per sample (colonna)
margin_height <- 200  # margine verticale extra (es. spazio per titolo, legenda, ecc.)
margin_width  <- 200  # margine orizzontale extra (es. spazio per labels)

cat(paste("\nThe length of the initial list of genes is:", len_initial_list, ", so considering the heatmaps are listing only the top", top_n, "up and downregulated, there are", (len_initial_list - 2*top_n), "genes not shown in the heatmaps", "\n"))
cat(paste("\nThe length of the HALLMARK list of genes is:", len_hallmark, ", so considering the heatmaps are listing only the top", top_n, "up and downregulated, there are", (len_hallmark - 2*top_n), "genes not shown in the heatmaps", "\n"))
cat(paste("\nThe length of the cytokines list of genes is:", len_cytokine, ", so considering the heatmaps are listing only the top", top_n, "up and downregulated, there are", (len_cytokine - 2*top_n), "genes not shown in the heatmaps", "\n"))
cat(paste("\nThe length of the inflammation list of genes is:", len_inflammation, ", so considering the heatmaps are listing only the top", top_n, "up and downregulated, there are", (len_inflammation - 2*top_n), "genes not shown in the heatmaps", "\n"))
cat(paste("\nThe length of the interferon list of genes is:", len_interferon, ", so the heatmap in this case is listing all the genes in that family", "\n"))
```

```{r all_genes_heatmaps, eval=FALSE}
# z_matrix_all_top <- z_matrix[genes_in_z_mostreg, ]

# genes_in_z_up <- head(intersect(all_genes_up, rownames(z_matrix)), 1000)
# z_matrix_all_topup <- z_matrix[genes_in_z_up, ]
# genes_in_z_down <- head(intersect(all_genes_down, rownames(z_matrix)), 1000)
# z_matrix_all_topdown <- z_matrix[genes_in_z_down, ]

# current_matrix <- z_matrix_all_top
# png(paste0(png_basepath, "/heatmap_allgenes_mostreg_byday.png"), 
#     width = (ncol(current_matrix) * cellwidth) + margin_width, 
#     height = (nrow(current_matrix) * cellheight) + margin_height)
# 
# pheatmap(
#   current_matrix,
#   cluster_rows = FALSE,
#   cluster_cols = FALSE,
#   show_rownames = TRUE,
#   show_colnames = TRUE,
#   cellheight = cellheight,
#   cellwidth = cellwidth,
#   color = colorRampPalette(c("blue", "white", "red"))(50),
#   main = paste("Top", top_n, "most regulated genes"),
#   legend = TRUE,
#   legend_breaks = c(-2, 0, 2),
#   legend_labels = c("-2 (Z-score)", "0", "2 (Z-score)")
# )
# 
# dev.off()
# 
# current_matrix <- z_matrix_all_topup
# png(paste0(png_basepath, "/heatmap_allgenes_mostreg_byday_up.png"), 
#     width = (ncol(current_matrix) * cellwidth) + margin_width, 
#     height = (nrow(current_matrix) * cellheight) + margin_height)
# 
# pheatmap(
#   current_matrix,
#   cluster_rows = FALSE,
#   cluster_cols = FALSE,
#   show_rownames = TRUE,
#   show_colnames = TRUE,
#   cellheight = cellheight,
#   cellwidth = cellwidth,
#   color = colorRampPalette(c("blue", "white", "red"))(50),
#   main = paste("Top", top_n, "most regulated genes"),
#   legend = TRUE,
#   legend_breaks = c(-2, 0, 2),
#   legend_labels = c("-2 (Z-score)", "0", "2 (Z-score)")
# )
# 
# dev.off()
# 
# current_matrix <- z_matrix_all_topdown
# png(paste0(png_basepath, "/heatmap_allgenes_mostreg_byday_down.png)", 
#     width = (ncol(current_matrix) * cellwidth) + margin_width, 
#     height = (nrow(current_matrix) * cellheight) + margin_height)
# 
# pheatmap(
#   current_matrix,
#   cluster_rows = FALSE,
#   cluster_cols = FALSE,
#   show_rownames = TRUE,
#   show_colnames = TRUE,
#   cellheight = cellheight,
#   cellwidth = cellwidth,
#   color = colorRampPalette(c("blue", "white", "red"))(50),
#   main = paste("Top", top_n, "most regulated genes"),
#   legend = TRUE,
#   legend_breaks = c(-2, 0, 2),
#   legend_labels = c("-2 (Z-score)", "0", "2 (Z-score)")
# )
# 
# dev.off()
```

```{r heatmap_if_saveas_png, eval=FALSE}
# current_matrix <- z_matrix_top_ordered_by_treatment
# png(paste0(png_basepath, "/heatmap_genelist_mostreg_bycondition.png"), 
#     width = (ncol(current_matrix) * cellwidth) + margin_width, 
#     height = (nrow(current_matrix) * cellheight) + margin_height)
# 
# pheatmap(
#   current_matrix,
#   cluster_rows = FALSE,
#   cluster_cols = FALSE,
#   show_rownames = TRUE,
#   show_colnames = TRUE,
#   cellheight = cellheight,
#   cellwidth = cellwidth,
#   color = colorRampPalette(c("blue", "white", "red"))(50),
#   main = paste("Top", top_n, "most regulated genes"),
#   legend = TRUE,
#   legend_breaks = c(-2, 0, 2),
#   legend_labels = c("-2 (Z-score)", "0", "2 (Z-score)")
# )
# 
# dev.off()
```

```{r generate_heatmaps_and_tables, results='hide'}

initial_genelists_matrices = list(
  list(z_matrix_top_ordered_by_day, "most regulated genes in the initial genelist,\nordered by day"),
  list(z_matrix_top_ordered_by_treatment, "most regulated genes in the initial genelist,\nordered by condition"),
  list(z_matrix_top_ordered_by_day_up, "most upregulated genes in the initial genelist,\nordered by day"),
  list(z_matrix_top_ordered_by_treatment_up, "most upregulated genes in the initial genelist,\nordered by condition"),
  list(z_matrix_top_ordered_by_day_down, "most downregulated genes in the initial genelist,\nordered by day"),
  list(z_matrix_top_ordered_by_treatment_down, "most downregulated genes in the initial genelist,\nordered by condition")
  )

heatmaps2plot <- list()
counter = 0

for (element in initial_genelists_matrices) {
  counter <- counter + 1
  matrix_data <- element[[1]]
  description <- element[[2]]
  
  heatmap_plot <- pheatmap(
    matrix_data,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    show_rownames = TRUE,
    show_colnames = TRUE,
    cellheight = cellheight,
    cellwidth = cellwidth,
    color = colorRampPalette(c("blue", "white", "red"))(50),
    main = paste("Top", top_n, description),
    legend = TRUE,
    legend_breaks = c(-2, 0, 2),
    legend_labels = c("-2 (Z-score)", "0", "2 (Z-score)")
  )
  
  # print(p)
  
  # Prendi i geni mostrati nella heatmap (rownames della matrice)
  genes_in_heatmap <- rownames(matrix_data)
  
  # Filtra df_combined per questi geni
  filtered_df <- df_combined[df_combined$gene %in% genes_in_heatmap, ]
  
  # Crea la datatable
  heat_table <- datatable(
    filtered_df,
    extensions = 'Buttons',
    options = list(
      dom = 'Blfrtip',
      buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
      lengthMenu = list(c(10, 25, 50, -1),
                        c(10, 25, 50, "All"))
    ),
    rownames = FALSE,
    caption = paste("Genes in heatmap:", description)
  )
  
  heatmaps2plot[[counter]] <- list(
    p = heatmap_plot,
    d = heat_table
  )
}
```

```{r plot_heatmaps_1, fig.height=11}
heatmaps2plot[[1]]$p
heatmaps2plot[[1]]$d
```

```{r plot_heatmaps_2, fig.height=11}
heatmaps2plot[[2]]$p
heatmaps2plot[[2]]$d
```

```{r plot_heatmaps_3, fig.height=11}
heatmaps2plot[[3]]$p
heatmaps2plot[[3]]$d
```

```{r plot_heatmaps_4, fig.height=11}
heatmaps2plot[[4]]$p
heatmaps2plot[[4]]$d
```

```{r plot_heatmaps_5, fig.height=11}
heatmaps2plot[[5]]$p
heatmaps2plot[[5]]$d
```

```{r plot_heatmaps_6, fig.height=11}
heatmaps2plot[[6]]$p
heatmaps2plot[[6]]$d
```

```{r testheat_2}
family_list <- list()

# Lista dei dati e dei nomi da assegnare
families <- list(
  hallmark    = list(df = df_genes_hallmark,       name = "HALLMARK_terms"),
  cytokine    = list(df = df_genes_go_cytokine,     name = "Cytokines_GO_term"),
  inflammation= list(df = df_genes_go_inflammation, name = "Inflammation_GO_term"),
  interferon  = list(df = df_genes_go_interferon,   name = "Interferon_GO_term")
)

# Popolamento di family_list
for (fam in names(families)) {
  family_list[[fam]] <- list(
    df = families[[fam]]$df,
    family_name = families[[fam]]$name
  )
}

# Lista per salvare matrici e info di plot
plots_to_do <- list()

for (family in names(family_list)) {
  
  df <- family_list[[family]]$df
  family_name <- family_list[[family]]$family_name
  
  if (!"log2FC_combined" %in% colnames(df)) {
    message(paste("Attenzione:", df, "non contiene la colonna log2FC_combined"))
    next
  }
  
  n_genes <- nrow(df)
  
  if (n_genes > top_n) {
    # Top upregulated
    df_sorted_up <- df[order(df$log2FC_combined, decreasing = TRUE), ]
    top_up_genes <- df_sorted_up$gene[1:top_n]
    matrix_up <- z_matrix[top_up_genes[top_up_genes %in% rownames(z_matrix)], , drop=FALSE]
    plots_to_do[[paste0(family, "_top", top_n, "_upregulated")]] <- list(
      matrix = matrix_up,
      title = paste(family_name, "- Top", top_n, "Upregulated Genes")
    )
    
    # Top downregulated
    df_sorted_down <- df[order(df$log2FC_combined, decreasing = FALSE), ]
    top_down_genes <- df_sorted_down$gene[1:top_n]
    matrix_down <- z_matrix[top_down_genes[top_down_genes %in% rownames(z_matrix)], , drop=FALSE]
    plots_to_do[[paste0(family, "_top", top_n, "_downregulated")]] <- list(
      matrix = matrix_down,
      title = paste(family_name, "- Top", top_n, "Downregulated Genes")
    )
    
  } else {
    # Meno di 50 geni: prendi tutti ordinati decrescente
    df_sorted <- df[order(df$log2FC_combined, decreasing = TRUE), ]
    all_genes <- df_sorted$gene
    matrix_all <- z_matrix[all_genes[all_genes %in% rownames(z_matrix)], , drop=FALSE]
    plots_to_do[[paste0(family, "_all_genes")]] <- list(
      matrix = matrix_all,
      title = paste(family_name, "- All Genes\nordered from upregulated to downregulated")
    )
  }
}

sample_order_by_time <- sample_df_by_time$sample
sample_order_by_cond <- sample_df_by_cond$sample

col_orders <- list(
  time_then_condition = sample_order_by_time,
  condition_then_time = sample_order_by_cond
)

heatmaps2plot <- list()
counter = 0

for (plot_name in names(plots_to_do)) {
  p <- plots_to_do[[plot_name]]
  mat <- p$matrix
  
  for (order_name in names(col_orders)) {
    counter <- counter + 1
    col_order <- col_orders[[order_name]]
    
    # Prendi solo le colonne che ci sono nella matrice e ordinale
    cols_present <- col_order[col_order %in% colnames(mat)]
    mat_ordered <- mat[, cols_present, drop=FALSE]
    
    plot <- pheatmap(mat_ordered,
                     cluster_rows = FALSE,
                     cluster_cols = FALSE,
                     show_rownames = TRUE,
                     show_colnames = TRUE,
                     cellheight = cellheight,
                     cellwidth = cellwidth,
                     color = colorRampPalette(c("blue", "white", "red"))(50),
                     main = paste(p$title, "\nColumns ordered by:", order_name),
                     legend = TRUE,
                     legend_breaks = c(-2, 0, 2),
                     legend_labels = c("-2 (Z-score)", "0", "2 (Z-score)")
    )
    
    # Prendi i geni (rownames) presenti nella matrice ordinata
    genes_in_heatmap <- rownames(mat_ordered)
    
    # Filtra df_combined per quei geni
    filtered_df <- df_combined[df_combined$gene %in% genes_in_heatmap, ]
    
    # Mostra la datatable con filtro
    res_dt <- datatable(
      filtered_df,
      extensions = 'Buttons',
      options = list(
        dom = 'Blfrtip',
        buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
        lengthMenu = list(c(10, 25, 50, -1),
                          c(10, 25, 50, "All"))
      ),
      rownames = FALSE,
      caption = paste("Genes in heatmap:", p$title, "\nColumns ordered by:", order_name)
    )
    
    heatmaps2plot[[counter]] <- list(
      p = plot,
      d = res_dt
    )
  }
}
```

```{r plot_heatmaps_12, fig.height=11}
heatmaps2plot[[1]]$p
heatmaps2plot[[1]]$d
```

```{r plot_heatmaps_22, fig.height=11}
heatmaps2plot[[2]]$p
heatmaps2plot[[2]]$d
```

```{r plot_heatmaps_32, fig.height=11}
heatmaps2plot[[3]]$p
heatmaps2plot[[3]]$d
```

```{r plot_heatmaps_42, fig.height=11}
heatmaps2plot[[4]]$p
heatmaps2plot[[4]]$d
```

```{r plot_heatmaps_52, fig.height=11}
heatmaps2plot[[5]]$p
heatmaps2plot[[5]]$d
```

```{r plot_heatmaps_62, fig.height=11}
heatmaps2plot[[6]]$p
heatmaps2plot[[6]]$d
```

```{r plot_heatmaps_72, fig.height=11}
heatmaps2plot[[7]]$p
heatmaps2plot[[7]]$d
```

```{r plot_heatmaps_82, fig.height=11}
heatmaps2plot[[8]]$p
heatmaps2plot[[8]]$d
```

```{r plot_heatmaps_92, fig.height=11}
heatmaps2plot[[9]]$p
heatmaps2plot[[9]]$d
```

```{r plot_heatmaps_102, fig.height=11}
heatmaps2plot[[10]]$p
heatmaps2plot[[10]]$d
```

```{r plot_heatmaps_112, fig.height=11}
heatmaps2plot[[11]]$p
heatmaps2plot[[11]]$d
```

```{r plot_heatmaps_122, fig.height=11}
heatmaps2plot[[12]]$p
heatmaps2plot[[12]]$d
```

```{r plot_heatmaps_132, fig.height=11}
heatmaps2plot[[13]]$p
heatmaps2plot[[13]]$d
```

```{r plot_heatmaps_142, fig.height=11}
heatmaps2plot[[14]]$p
heatmaps2plot[[14]]$d
```

<div style="color:red;">

##### Gene expression Heatmap

ongoing

</div>

## Library Sizes

A bar plot displaying total read counts per sample is shown below

```{r counts_dds, results="asis", fig.width=8}

lib_sizes <- data.frame(
  sample_description = colData(dds)$sample,
  total_counts = colSums(counts(dds))
)

# Plot
ggplot(lib_sizes, aes(x = sample_description, y = total_counts)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "",
       x = "Sample",
       y = "Total Counts") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

<div style="color:red;">

### Interpretation Library Sizes

Variable but not extremely variable

</div>

## Violin plot of VST-normalized counts

Below we present a violin plot of the VST-normalized read counts by sample.

A violin plot of VST-normalized counts provides an overview of the global distribution of gene expression values across samples after normalization.  This plot allows for the detection of potential outliers, technical biases, or inconsistencies in distribution across samples, which could affect downstream analyses. A consistent distribution of VST counts across samples suggests successful normalization and comparable expression profiles.

```{r violin_dds, results="asis", fig.width=8, fig.height=5}

rld_matrix <- assay(rld)
df_long <- reshape2::melt(rld_matrix, varnames = c("gene", "sample"), value.name = "expression")

ggplot(df_long, aes(x = sample, y = expression)) +
  geom_violin(trim = FALSE, fill = "#8da0cb", color = "black") +
  geom_boxplot(width = 0.1, outlier.size = 0.5) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "VST-normalized expression",
    title = ""
  )
```

<div style="color:red;">

### Interpretation Violin plot of VST-normalized counts

The violin plot of counts data displays a consistent distribution of VST counts across samples.  
This indicates no substantial differences in gene expression profiles between the conditions and confirms the quality and reliability of the samples, supporting the inclusion of all samples in subsequent analyses.

</div>

### IFI6 counts

Below we present the distribution of IFI6 normalised expression across conditions

```{r}
ifi6 = df_long[df_long$gene == "IFI6", ]
ifi6$group <- unlist(lapply(strsplit(as.character(ifi6$sample), "_D"), "[[", 1))

ggplot(ifi6, aes(x = group, y = expression)) +
  geom_violin(trim = FALSE, fill = "#8da0cb", color = "black") +
  geom_boxplot(width = 0.1, outlier.size = 0.5) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "VST-normalized expression",
    title = ""
  )
```

## Average Expression across all samples

Below we present the distributions of the selected genes expression across conditions in a barplot where the height of each bar represents the average expression of all the selected genes in that particular condition. Each condition includes all the available replicates except those already filtered. 

```{r load_data, results='asis'}
counts_matrix <- assay(dds, "counts")  # raw counts
vst_matrix <- assay(dds, "vst") 
  
for (i in params$genelists) {
    
  list <- read.csv(i)[[1]]
  
  cat(paste("\n\n### Average Expression, Raw counts, gene list:", basename(i), "\n\n"))
  
  count_matrix_subset <- counts_matrix[rownames(counts_matrix) %in% list, ]
  
  conditions <- gsub("_rep[0-9]+$", "", colnames(count_matrix_subset))
  t_mat <- t(count_matrix_subset)
  summed_mat <- rowsum(t_mat, group = conditions)
  rep_counts <- table(conditions)
  mean_mat <- sweep(summed_mat, 1, rep_counts, FUN = "/")
  mean_mat <- t(mean_mat)
  mean_expr_per_sample <- colMeans(mean_mat)
  df <- data.frame(
    sample = names(mean_expr_per_sample),
    mean_expression = mean_expr_per_sample
  )
  
  # plot con ggplot2
  plot1 <- ggplot(df, aes(x = sample, y = mean_expression)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    xlab("Condition") +
    ylab("Raw counts average gene expression") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(plot1)
  
  cat(paste("\n\n### Average Expression, VST normalized counts, gene list:", basename(i), "\n\n"))
  
  vst_matrix_subset <- vst_matrix[rownames(vst_matrix) %in% list, ]
  
  conditions <- gsub("_rep[0-9]+$", "", colnames(vst_matrix_subset))
  t_mat <- t(vst_matrix_subset)
  summed_mat <- rowsum(t_mat, group = conditions)
  rep_counts <- table(conditions)
  mean_mat <- sweep(summed_mat, 1, rep_counts, FUN = "/")
  mean_mat <- t(mean_mat)
  mean_expr_per_sample <- colMeans(mean_mat)
  df <- data.frame(
    sample = names(mean_expr_per_sample),
    mean_expression = mean_expr_per_sample
  )
  
  # plot con ggplot2
  plot2 <- ggplot(df, aes(x = sample, y = mean_expression)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    theme_minimal() +
    xlab("Condition") +
    ylab("VST normalized average gene expression") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(plot2)
}
```

